Fast Sorting on Modern Computers

### 2005-07-30 docs/tech_report/background.tex
minor changes to references
fixing up lots of things

\section{Big O notation}
Big O notation is used to describe the complexity of an algorithm. It says that
the algorithm will complete in a certain number of steps, in relation to number
of items of data being worked on, without giving specifics of how many
instructions each step will take to execute. It is possible, therefore, that an
algorithm with a lower complexity will complete slower than an algorithm where
each step is executed very quickly, but which has a greater order of complexity.

@@

An algorithm which is $O(N)$ is said to complete in linear time. An algorithm
which is $O(N^2)$ is said to complete in quadratic time.  $O(NlogN)$ algorithms
may take several orders of magnitude more time than $O(N)$ algorithms, and
typically use the \n{divide-and-conquer} principle, or use a binary tree to
reduce complexity. Quicksort, mergesort and heapsort are all of this type,
whereas elementary sorts, such as insertion sort, selection sort, bubblesort and
shakersort are $O(N^2)$ algorithms. Radixsort is $O(N)$, but can be more
expensive than quicksort, due to the cost of each step. These are discussed in
later chapters.

Not all algorithms have an easily derivable complexity. Many studies have
attempted to tie down the complexity of shellsort, and it is not easily
definable in terms of the catageories above (see \ref{TODO}).

>>>>

\section{Big O Notation}
Big O notation is used to describe the complexity of an algorithm. It says that
the algorithm will complete in a certain number of steps, in relation to number
of items of data being worked on, without giving specifics of how many
instructions each step will take to execute. It is possible, therefore, that an
algorithm with a lower complexity will complete slower than an algorithm where
each step is executed very quickly, but which has a greater order of complexity.

@@

An algorithm which is $O(N)$ is said to complete in linear time. An algorithm
which is $O(N^2)$ is said to complete in quadratic time.  $O(NlogN)$ algorithms
may take several orders of magnitude more time than $O(N)$ algorithms, and
typically use the \n{divide-and-conquer} principle, or use a binary tree to
reduce complexity. Quicksort, mergesort and heapsort are all of this type,
whereas elementary sorts, such as insertion sort, selection sort, bubblesort and
shakersort are $O(N^2)$ algorithms. Radixsort is $O(N)$, but can be more
expensive than quicksort, due to the cost of each step. These are discussed in
later chapters. The shellsort we use, called \n{Gonnet Shellsort} has a
complexity of TODO (see \ref{TODO}).

### 2005-07-30 docs/tech_report/conclusions.tex
fixing up lots of things

\afterpage{
\thispagestyle{empty}
\clearpage
\enlargethispage{14em}
\vspace*{-10em}
\setcounter{figure}{8}
\begin{figure}[H]
\begin{changemargin}
\includegraphics[angle=90,scale=1.8]{plots/all_cycles}
\vspace*{8em}
\end{changemargin}
\caption{\label{All cycles}Cycles per key of several major sorts and their variations - this was
measured on a Pentium 4 using hardware performance counters.}
\end{figure}
\newpage
}

>>>>

\afterpage{
\thispagestyle{empty}
\clearpage
\enlargethispage{14em}
\vspace*{-10em}
\setcounter{figure}{8}
\begin{figure}[H]
\begin{changemargin}
\includegraphics[angle=90,scale=1.8]{plots/all_cycles}
\vspace*{8em}
\end{changemargin}
\caption[Cycles per key of several major sorts and their variations]{\label{All cycles}Cycles per key of several major sorts and their variations - this was
measured on a Pentium 4 using hardware performance counters.}
\end{figure}
\newpage
}

### 2005-07-30 docs/tech_report/heap.tex
fixing up lots of things

\subsection{Implicit heaps}

@@

\subsection{Sorting with heaps}
To sort an array with a heap, it is necessary to first create the heap. This is
known as \n{heapifying}, and can be performed in two ways.

@@

\subsubsection{In-place sorting}
For an in-place sort, a heap is created as specified above. Then the root of the
heap (the largest key) is swapped with the last key in the heap, and the new
root is sifted down the heap which has been shrunk so as not to include the old
root, which is now in its final position. This is repeated for the entire heap,
leaving a fully sorted array.

>>>>

\subsection{Implicit Heaps}

@@

\subsection{Sorting with Heaps}
To sort an array with a heap, it is necessary to first create the heap. This is
known as \n{heapifying}, and can be performed in two ways.

@@

\subsubsection{In-place Sorting}
For an in-place sort, a heap is created as specified above. Then the root of the
heap (the largest key) is swapped with the last key in the heap, and the new
root is sifted down the heap which has been shrunk so as not to include the old
root, which is now in its final position. This is repeated for the entire heap,
leaving a fully sorted array.

### 2005-07-30 docs/tech_report/method.tex
fixing up lots of things

\section{Implementation of sorts}

@@

\section{Future work}
An improvement that would have yielded interesting results would be to do simulations
which only consisted of flow control, with no comparisons. This would allow a
division of flow control misses and comparative misses. However, this would not
always be possible; sorts like quicksort and heapsort use their data to
control the flow of the program. In addition, bubblesort, shakersort and
selection sort have (analytically) predictable flow control, while radixsort has
no comparison misses at all. Mergesort may benefit from this, however, especially versions
with difficult flow control diagrams, such as \n{algorithm N} and \n{algorithm
S} (see Section \ref{Algorithm N}).

>>>>

\section{Sort Implementation}

@@

\section{Future Work}
An improvement that would have yielded interesting results would be to do simulations
which only consisted of flow control, with no comparisons. This would allow a
division of flow control misses and comparative misses. However, this would not
always be possible; sorts like quicksort and heapsort use their data to
control the flow of the program. In addition, bubblesort, shakersort and
selection sort have (analytically) predictable flow control, while radixsort has
no comparison misses at all. Mergesort may benefit from this, however, especially versions
with difficult flow control diagrams, such as \n{algorithm N} and \n{algorithm
S} (see Section \ref{Algorithm N}).

### 2005-07-30 docs/tech_report/report.tex
made the thing compile now
reread the merge - updated results and created expected vs actual added a label to heap added a label to smallplot in report change multi mergesort to multi-mergesort
minor changes to references
fixing up lots of things

\newcommand{\smallscale}{.74}
\newcommand{\smallplot}[2]{
\setcounter{subfigure}{0}
\thispagestyle{empty}
\clearpage
\enlargethispage{14em}
\vspace*{-5em}
\begin{figure}[H]
\begin{changemargin}
\subfigure[Instructions per key] {\includegraphics[scale=0.74]{plots/#1_-_Instructions.eps}}\\
\subfigure[Level 1 cache misses per key]	{\includegraphics[scale=0.74]{plots/#1_-_Level_1_misses.eps}}
\subfigure[Level 2 cache misses per key]	{\includegraphics[scale=0.74]{plots/#1_-_Level_2_misses.eps}}
\subfigure[Branches per key]				{\includegraphics[scale=0.74]{plots/#1_-_Branches.eps}}
\subfigure[Branch misses per key]			{\includegraphics[scale=0.74]{plots/#1_-_Branch_misses.eps}}
\end{changemargin}
\vspace*{6em}
\caption{Simulation results for #2}
\label{Simulation results for #2}
\end{figure}
\newpage
}

\newcommand{\sixplot}[7]{
\afterpage{
\setcounter{subfigure}{0}
\thispagestyle{empty}
\clearpage
\enlargethispage{14em}
\vspace*{-5em}
\begin{figure}[H]
\begin{changemargin}
\subfigure[]{\label{#7-1}\includegraphics[scale=0.74]{../../counters/#1}}
\ifthenelse{\equal{#2}{}}{}{\subfigure[]{\label{#7-2}\includegraphics[scale=0.74]{../counters/#2}}}
\ifthenelse{\equal{#3}{}}{}{\subfigure[]{\label{#7-3}\includegraphics[scale=0.74]{../counters/#3}}}
\ifthenelse{\equal{#4}{}}{}{\subfigure[]{\label{#7-4}\includegraphics[scale=0.74]{../counters/#4}}}
\ifthenelse{\equal{#5}{}}{}{\subfigure[]{\label{#7-5}\includegraphics[scale=0.74]{../counters/#5}}}
\ifthenelse{\equal{#6}{}}{}{\subfigure[]{\label{#7-6}\includegraphics[scale=0.74]{../counters/#6}}}
\end{changemargin}
\vspace*{6em}
\caption{Branch prediction performance for #7. All graphs use the same scale.}
\label{Branch prediction performance for #7}
\end{figure}
\newpage
}}

>>>>

\newcommand{\smallscale}{.74}
\newcommand{\smallplot}[2]{
\setcounter{subfigure}{0}
\thispagestyle{empty}
\clearpage
\enlargethispage{14em}
\vspace*{-5em}
\begin{figure}[H]
\begin{changemargin}
\subfigure[Cycles per key] {\includegraphics[scale=0.74]{plots/#1-_cycles.eps}}
\subfigure[Instructions per key] {\includegraphics[scale=0.74]{plots/#1_-_Instructions.eps}}
\subfigure[Level 1 cache misses per key]	{\includegraphics[scale=0.74]{plots/#1_-_Level_1_misses.eps}}
\subfigure[Level 2 cache misses per key]	{\label{#2 level 2 misses}\includegraphics[scale=0.74]{plots/#1_-_Level_2_misses.eps}}
\subfigure[Branches per key]				{\includegraphics[scale=0.74]{plots/#1_-_Branches.eps}}
\subfigure[Branch misses per key]			{\includegraphics[scale=0.74]{plots/#1_-_Branch_misses.eps}}
\end{changemargin}
\vspace*{6em}
\caption{Simulation results for #2}
\label{Simulation results for #2}
\end{figure}
\newpage
}

\newcommand{\sixplot}[7]{
\afterpage{
\setcounter{subfigure}{0}
\thispagestyle{empty}
\clearpage
\enlargethispage{14em}
\vspace*{-5em}
\begin{figure}[H]
\begin{changemargin}
\subfigure[]{\label{#7-1}\includegraphics[scale=0.74]{plots/counter_#1}}
\ifthenelse{\equal{#2}{}}{}{\subfigure[]{\label{#7-2}\includegraphics[scale=0.74]{plots/counter_#2}}}
\ifthenelse{\equal{#3}{}}{}{\subfigure[]{\label{#7-3}\includegraphics[scale=0.74]{plots/counter_#3}}}
\ifthenelse{\equal{#4}{}}{}{\subfigure[]{\label{#7-4}\includegraphics[scale=0.74]{plots/counter_#4}}}
\ifthenelse{\equal{#5}{}}{}{\subfigure[]{\label{#7-5}\includegraphics[scale=0.74]{plots/counter_#5}}}
\ifthenelse{\equal{#6}{}}{}{\subfigure[]{\label{#7-6}\includegraphics[scale=0.74]{plots/counter_#6}}}
\end{changemargin}
\vspace*{6em}
\caption[Branch prediction performance for #7]{Branch prediction performance for #7. All graphs use the same scale.}
\label{Branch prediction performance for #7}
\end{figure}
\newpage
}}
